---
title: ""
format:
  pdf:
    documentclass: article
    toc: false
    number-sections: true
    fig-cap-location: bottom
    papersize: a4
    fontsize: 11pt
    geometry:
      - margin=2.2cm
    include-in-header:
      text: |
        \usepackage{fancyhdr}
        \usepackage{setspace}
        \usepackage{tocloft}
        \pagestyle{fancy}
        \fancyhf{}
        \rhead{\thepage}
        \renewcommand{\headrulewidth}{0pt}
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

\thispagestyle{empty}

\vspace*{3cm}

\begin{center}
{\Large \textbf{Housing Lending, Overseas Migration and Wage Growth Across Australian States}}\\[0.8cm]

{\large Data Exploration Project}\\[1.8cm]

TingTing Wu\\
Student ID: 35717343\\[1.2cm]

FIT5147 Data Exploration and Visualization\\
Semester 1, 2026\\
Monash University\\[1.2cm]

29 March 2026
\end{center}

\newpage

\thispagestyle{plain}
\begin{center}
\section*{Contents}
\end{center}

\tableofcontents

\newpage

# Introduction

This report, *Housing Lending, Overseas Migration and Wage Growth Across Australian States*, examines one of the most widely debated social and economic issues in Australia: housing affordability. Recent public discussion has frequently linked rising housing pressure to migration, limited housing supply, and weak wage growth. For example, SBS News discussed how high migration is often blamed for the housing crisis and questioned whether this explanation is supported by evidence (SBS News, 2026). This makes the topic highly topical and important for students, workers, households, and policy makers.

The main purpose of this project is to explore how housing lending, overseas migration, and wage growth vary across Australian states over time. Rather than focusing only on house prices, this report uses several Australian Bureau of Statistics datasets to examine broader patterns in owner-occupier lending, investor lending, net overseas migration, and the Wage Price Index. The problem explored in this report is whether common public claims about migration, wages, and housing stress are reflected in the data, and whether these patterns differ across states.

I am personally motivated to study this topic because, as an international student from Taiwan living in Melbourne, I have become more aware of how difficult housing affordability has become for students and workers. I have seen how high housing costs can make long-term financial security feel increasingly out of reach for ordinary people. Public debate in Melbourne has also intensified around whether migration is a major cause of housing pressure. This motivated me to examine the issue more carefully using data rather than relying only on media opinions or political claims.

This report addresses the following questions:

1.  Does the growth of housing debt align with wage growth across Australian states, and what is the current debt structure?
2.  How does the regional distribution of immigrants relate to local wages and housing costs?
3.  Is investor lending activity driven by migration demand or market speculation?

# Data Wrangling and Checking

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(lubridate)
library(knitr)
library(kableExtra)
library(forcats)
```

This project uses four open datasets from the Australian Bureau of Statistics (ABS): Net Overseas Migration by Country of Birth, Owner-Occupier Lending, Investor Lending, and the Wage Price Index. The direct download links are provided below:

-   [Net Overseas Migration by Country of Birth](https://www.abs.gov.au/statistics/people/population/overseas-migration/latest-release/34070DO001_202425.xlsx)
-   [Owner-Occupier Lending](https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release/560104.xlsx)
-   [Investor Lending](https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release/560114.xlsx)
-   [Wage Price Index](https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/wage-price-index-australia/latest-release/634502b.xlsx)

Before checking and cleaning the data, I first transformed the three ABS datasets into tidy formats that were easier to compare and analyse in R. Although all three sources were downloaded as Excel workbooks, they had different internal structures and could not be used directly for analysis.

For the overseas migration data, the workbook contained nine separate tables, including one national table and eight state or territory tables. I created a lookup table containing the sheet names, state names, and row ranges, then used a function to read each worksheet in the same way. In each sheet, I used the first imported row as the column names, renamed the first two columns as `sacc_code` and `country_of_birth`, and removed note labels from the year columns, which is the letter attached to 2024-25. I then converted the data from wide format to long format using `pivot_longer()`, so that each row represented one state, one country of birth, and one financial year. Finally, I combined all nine sheets into a single dataset called `overseas_migration`.

For the housing finance data, I used the `Data1` sheet from both the owner-occupier and investor lending workbooks. In both files, the first ten rows contained metadata rather than observations, so I read these rows separately from the main time series data. I created temporary column names for the series columns and extracted key metadata such as state, value type, unit, and series type by splitting the original ABS series description. I then reshaped the time series data into long format and created additional time variables, including calendar year, quarter, and financial year. After that, I combined the owner-occupier and investor datasets into one dataset called `housing_finance`, with an added variable called `loan_type` to distinguish between the two sources.

For the Wage Price Index data, I again used the `Data1` sheet and followed a similar process. I read the metadata rows separately, created temporary series column names, and extracted information such as state, wage measure, and unit from the ABS series descriptions. I then reshaped the data into long format and created year, quarter, and financial year variables. This produced a final tidy dataset called `wage_price_index`.

Overall, the main purpose of this step was to convert the original Excel workbooks into consistent long-format datasets that could be compared more easily across states and time. This also made it easier to carry out the next stage of the project, which involved checking the structure, identifying missing values, reviewing inconsistencies, and cleaning the data where necessary. The tidy datasets and all data processing code used in this project are available in my [GitHub repository](https://github.com/ruby910113/FIT5147-DEP-report).

```{r}
# File path for the overseas migration workbook
om_file_path <- "data/34070DO001_202425.xlsx"

# Sheet names, state names, and row ranges
om_sheet_info <- tibble(
  sheet = paste0("Table 1.", 1:9),
  state = c("Australia", "New South Wales", "Victoria", "Queensland", "South Australia", "Western Australia", "Tasmania", "Northern Territory", "Australian Capital Territory"),
  range = c("A14:W262", rep("A13:W261", 8)))

# Read one overseas migration sheet and tidy it
read_om_sheet <- function(sheet, state, range, file_path) {
  
  om_raw <- read_excel(path = file_path, sheet = sheet, range = range, col_names = FALSE)
  
  # Use the first row as column names
  om_col_names <- om_raw[1, ] %>%
    unlist() %>%
    as.character()
  
  om_data <- om_raw[-1, ]
  names(om_data) <- om_col_names
  
  # Rename the first two columns
  names(om_data)[1:2] <- c("sacc_code", "country_of_birth")
  
  # Remove notes (such as (d)) from the year columns
  om_year_cols <- names(om_data)[-(1:2)]
  om_clean_years <- str_remove(om_year_cols, "\\([A-Za-z]+\\)$")
  names(om_data)[-(1:2)] <- om_clean_years
  
  om_data %>%
    mutate(state = state,
           sacc_code = str_pad(as.character(sacc_code), width = 4, pad = "0")) %>%
    pivot_longer(cols = matches("^\\d{4}-\\d{2}$"),
                 names_to = "financial_year",
                 values_to = "net_overseas_migration") %>%
    mutate(net_overseas_migration = as.numeric(net_overseas_migration)) %>%
    select(state, sacc_code, country_of_birth, financial_year, net_overseas_migration)
}

# Combine all sheets into one dataset
overseas_migration <- pmap_dfr(om_sheet_info, \(sheet, state, range) read_om_sheet(sheet, state, range, om_file_path)) %>%
  filter(!is.na(country_of_birth))

rm(om_sheet_info)
```

```{r}
# Read one housing finance workbook from Data1
read_housing_finance_data1 <- function(file_path, loan_type) {
  
  # Read the metadata rows
  hf_meta <- read_excel(path = file_path, sheet = "Data1", n_max = 10, col_names = FALSE)
  
  # Read the time series data
  hf_data <- read_excel(path = file_path, sheet = "Data1", skip = 10, col_names = FALSE, col_types = c("date", rep("numeric", ncol(hf_meta) - 1)))
  
  # Create temporary column names
  series_cols <- paste0("series_", seq_len(ncol(hf_data) - 1))
  
  names(hf_meta) <- c("label", series_cols)
  names(hf_data) <- c("date", series_cols)
  
  # Make a table for the metadata
  hf_series_info <- tibble(
    series_col = series_cols,
    full_name = as.character(unlist(hf_meta[1, -1])),
    unit = as.character(unlist(hf_meta[2, -1])),
    series_type = as.character(unlist(hf_meta[3, -1])),
    data_type = as.character(unlist(hf_meta[4, -1])),
    frequency = as.character(unlist(hf_meta[5, -1]))) %>%
    separate(full_name,
             into = c("sector", "collection", "borrower_type", "dwelling_type", "state", "measure", "value_type", "blank"),
             sep = "\\s*;\\s*",
             extra = "drop",
             fill = "right") %>%
    select(series_col, state, borrower_type, dwelling_type, measure, value_type, unit, series_type)
  
  # Turn the data into long format
  hf_long <- hf_data %>%
    mutate(year = year(date),
           month = month(date),
           financial_year = if_else(
             month(date) >= 7,
             paste0(year(date), "-", str_sub(year(date) + 1, 3, 4)),
             paste0(year(date) - 1, "-", str_sub(year(date), 3, 4)))) %>%
    pivot_longer(cols = starts_with("series_"),
                 names_to = "series_col",
                 values_to = "housing_finance") %>%
    left_join(hf_series_info, by = "series_col") %>%
    mutate(loan_type = loan_type) %>%
    select(year, month, financial_year, state, loan_type, value_type, unit, series_type, housing_finance)
  
  hf_long
}

# Combine owner-occupier and investor data
housing_finance <- bind_rows(
  read_housing_finance_data1(
    file_path = "data/560104.xlsx", loan_type = "owner_occupier"),
  read_housing_finance_data1(
    file_path = "data/560114.xlsx", loan_type = "investor"))
```

```{r}
# File path for the wage price index workbook
wpi_file_path <- "data/634502b.xlsx"

# Read the metadata rows
wpi_meta <- read_excel(path = wpi_file_path, sheet = "Data1", n_max = 10, col_names = FALSE)

# Read the actual data
wpi_data <- read_excel(path = wpi_file_path, sheet = "Data1", skip = 10, col_names = FALSE, col_types = c("date", rep("numeric", ncol(wpi_meta) - 1)))

# Create temporary column names
series_cols <- paste0("series_", seq_len(ncol(wpi_data) - 1))

names(wpi_meta) <- c("label", series_cols)
names(wpi_data) <- c("date", series_cols)

# Make a metadata table
wpi_series_info <- tibble(
  series_col = series_cols,
  full_name = as.character(unlist(wpi_meta[1, -1])),
  unit = as.character(unlist(wpi_meta[2, -1]))) %>%
  separate(full_name, into = c("wpi_measure", "pay_type", "state", "sector", "industry", "blank"), sep = "\\s*;\\s*", extra = "drop", fill = "right") %>%
  mutate(
    across(c(wpi_measure, pay_type, state, sector, industry), str_squish),
    wpi_measure = case_when(
      wpi_measure == "Quarterly Index" ~ "index",
      wpi_measure == "Percentage Change from Previous Quarter" ~ "pct_change_prev_quarter",
      wpi_measure == "Percentage Change from Corresponding Quarter of Previous Year" ~ "pct_change_prev_year",
      TRUE ~ wpi_measure)) %>%
  select(series_col, state, wpi_measure, unit)

# Turn the data into long format
wage_price_index <- wpi_data %>%
  mutate(year = year(date), month = month(date), 
         financial_year = if_else(
           month(date) >= 7,
           paste0(year(date), "-", str_sub(year(date) + 1, 3, 4)),
           paste0(year(date) - 1, "-", str_sub(year(date), 3, 4)))) %>%
  pivot_longer(cols = starts_with("series_"), names_to = "series_col", values_to = "wage_price_index") %>%
  left_join(wpi_series_info, by = "series_col") %>%
  select(year, month, financial_year, state, wpi_measure, unit, wage_price_index)

rm(wpi_data, wpi_meta, wpi_series_info)
```

## Creating a Common Financial Year Structure

As shown in @tbl-time-summary, Housing finance and Wage price index are quarterly datasets recorded using calendar year and month, whereas Overseas migration is recorded by financial year. To make the three datasets more comparable, I created a financial year variable for the quarterly datasets based on the quarter-ending months of September, December, March, and June.

```{r}
#| label: tbl-time-summury
#| tbl-cap: Summary of time structure across the datasets

data_time_summary <- bind_rows(
  tibble(
    dataset = "Overseas migration",
    start_ym = NA_character_,
    end_ym = NA_character_,
    frequency = "Annual",
    time_period = paste0(
      min(overseas_migration$financial_year), " to ",
      max(overseas_migration$financial_year))),
  
  housing_finance %>%
    distinct(year, month, financial_year) %>%
    mutate(date = make_date(year, month, 1)) %>%
    summarise(
      dataset = "Housing finance",
      start_ym = paste0(year(min(date)), "/", month(min(date))),
      end_ym = paste0(year(max(date)), "/", month(max(date))),
      frequency = "Quarterly",
      time_period = paste0(
        min(financial_year), " to ",
        max(financial_year))),
  
  wage_price_index %>%
    distinct(year, month, financial_year) %>%
    mutate(date = make_date(year, month, 1)) %>%
    summarise(
      dataset = "Wage price index",
      start_ym = paste0(year(min(date)), "/", month(min(date))),
      end_ym = paste0(year(max(date)), "/", month(max(date))),
      frequency = "Quarterly",
      time_period = paste0(
        min(financial_year), " to ",
        max(financial_year))))

data_time_summary %>%
  kable(
    caption = "Summary of time structure across the datasets",
    col.names = c("Dataset", "Start (Year/Month)", "End (Year/Month)", "Frequency", "Time period")) %>%
  kable_styling(full_width = FALSE)
```

As shown in @fig-time-coverage, the three datasets have different time ranges. Housing finance and Wage price index start earlier than Overseas migration, so the full periods do not fully match. To make the comparison fairer, I focused on the common comparison period shown in grey, which is the overlapping financial-year range shared by all three datasets.

```{r}
#| label: fig-time-coverage
#| fig-cap: Financial-year coverage and common comparison period across the datasets

period_df <- bind_rows(
  overseas_migration %>%
    distinct(financial_year) %>%
    summarise(dataset = "Overseas migration", start_fy = min(financial_year), end_fy = max(financial_year)),
  
  housing_finance %>%
    distinct(financial_year) %>%
    summarise(dataset = "Housing finance", start_fy = min(financial_year), end_fy = max(financial_year)),
  
  wage_price_index %>%
    distinct(financial_year) %>%
    summarise(dataset = "Wage price index", start_fy = min(financial_year), end_fy = max(financial_year))) %>%
  mutate(start_num = as.numeric(str_sub(start_fy, 1, 4)), end_num = as.numeric(str_sub(end_fy, 1, 4)))

overlap_start <- max(period_df$start_num)
overlap_end <- min(period_df$end_num)

fy_breaks <- seq(floor(min(period_df$start_num)), ceiling(max(period_df$end_num)), by = 2)

fy_labels <- paste0(
  fy_breaks, "-",
  str_sub(fy_breaks + 1, 3, 4))

ggplot(period_df, aes(y = fct_rev(dataset))) +
  annotate("rect", xmin = overlap_start, xmax = overlap_end, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.5) +
  geom_segment(
    aes(x = start_num, xend = end_num, yend = fct_rev(dataset), colour = dataset),
    linewidth = 1.4) +
  geom_point(aes(x = start_num, colour = dataset), size = 3) +
  geom_point(aes(x = end_num, colour = dataset), size = 3) +
  annotate("text", x = (overlap_start + overlap_end) / 2, y = 3.35, label = "Grey area: common comparison period", size = 3.8,
    colour = "grey20") +
  scale_x_continuous(breaks = fy_breaks, labels = fy_labels, expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Financial-year coverage across the datasets", x = "Financial year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(), legend.position = "none")
```


# Data Exploration

# Conclusion

# Reflection

# References
