---
title: ""
format:
  pdf:
    documentclass: article
    toc: false
    number-sections: true
    fig-cap-location: bottom
    papersize: a4
    fontsize: 11pt
    geometry:
      - margin=2.2cm
    include-in-header:
      text: |
        \usepackage{fancyhdr}
        \usepackage{setspace}
        \usepackage{tocloft}
        \pagestyle{fancy}
        \fancyhf{}
        \rhead{\thepage}
        \renewcommand{\headrulewidth}{0pt}
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

\thispagestyle{empty}

\vspace*{3cm}

\begin{center}
{\Large \textbf{Housing Lending, Overseas Migration and Wage Growth Across Australian States}}\\[0.8cm]

{\large Data Exploration Project}\\[1.8cm]

TingTing Wu\\
Student ID: 35717343\\[1.2cm]

FIT5147 Data Exploration and Visualization\\
Semester 1, 2026\\
Monash University\\[1.2cm]

29 March 2026
\end{center}

\newpage

\thispagestyle{plain}
\begin{center}
\section*{Contents}
\end{center}

\tableofcontents

\newpage

# Introduction

This report, *Housing Lending, Overseas Migration and Wage Growth Across Australian States*, examines one of the most widely debated social and economic issues in Australia: housing affordability. Recent public discussion has frequently linked rising housing pressure to migration, limited housing supply, and weak wage growth. For example, SBS News discussed how high migration is often blamed for the housing crisis and questioned whether this explanation is supported by evidence (SBS News, 2026). This makes the topic highly topical and important for students, workers, households, and policy makers.

The main purpose of this project is to explore how housing lending, overseas migration, and wage growth vary across Australian states over time. Rather than focusing only on house prices, this report uses several Australian Bureau of Statistics datasets to examine broader patterns in owner-occupier lending, investor lending, net overseas migration, and the Wage Price Index. The problem explored in this report is whether common public claims about migration, wages, and housing stress are reflected in the data, and whether these patterns differ across states.

I am personally motivated to study this topic because, as an international student from Taiwan living in Melbourne, I have become more aware of how difficult housing affordability has become for students and workers. I have seen how high housing costs can make long-term financial security feel increasingly out of reach for ordinary people. Public debate in Melbourne has also intensified around whether migration is a major cause of housing pressure. This motivated me to examine the issue more carefully using data rather than relying only on media opinions or political claims.

This report addresses the following questions:

1.  Does the growth of housing lending align with wage growth across Australian states, and how has the lending structure changed over time?
2.  How does the regional distribution of immigrants relate to local wages and housing costs?
3.  Is investor lending activity driven by migration demand or market speculation?

# Data Wrangling and Checking

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(lubridate)
library(knitr)
library(kableExtra)
library(forcats)
library(ggbeeswarm)
library(patchwork)
library(scales)
library(slider)
```

This project uses four open datasets from the Australian Bureau of Statistics (ABS): Net Overseas Migration by Country of Birth, Owner-Occupier Lending, Investor Lending, and the Wage Price Index. The direct download links are provided below:

-   [Net Overseas Migration by Country of Birth](https://www.abs.gov.au/statistics/people/population/overseas-migration/latest-release/34070DO001_202425.xlsx)
-   [Owner-Occupier Lending](https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release/560104.xlsx)
-   [Investor Lending](https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release/560114.xlsx)
-   [Wage Price Index](https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/wage-price-index-australia/latest-release/634502b.xlsx)

Before checking and cleaning the data, I first transformed the three ABS datasets into tidy formats that were easier to compare and analyse in R. Although all three sources were downloaded as Excel workbooks, they had different internal structures and could not be used directly for analysis.

For the overseas migration data, the workbook contained nine separate tables, including one national table and eight state or territory tables. I created a lookup table containing the sheet names, state names, and row ranges, then used a function to read each worksheet in the same way. In each sheet, I used the first imported row as the column names, renamed the first two columns as `sacc_code` and `country_of_birth`, and removed note labels from the year columns, which is the letter attached to 2024-25. I then converted the data from wide format to long format using `pivot_longer()`, so that each row represented one state, one country of birth, and one financial year. Finally, I combined all nine sheets into a single dataset called `overseas_migration`.

For the housing finance data, I used the `Data1` sheet from both the owner-occupier and investor lending workbooks. In both files, the first ten rows contained metadata rather than observations, so I read these rows separately from the main time series data. I created temporary column names for the series columns and extracted key metadata such as state, value type, unit, and series type by splitting the original ABS series description. I then reshaped the time series data into long format and created additional time variables, including calendar year, quarter, and financial year. After that, I combined the owner-occupier and investor datasets into one dataset called `housing_finance`, with an added variable called `loan_type` to distinguish between the two sources.

For the Wage Price Index data, I again used the `Data1` sheet and followed a similar process. I read the metadata rows separately, created temporary series column names, and extracted information such as state, wage measure, and unit from the ABS series descriptions. I then reshaped the data into long format and created year, quarter, and financial year variables. This produced a final tidy dataset called `wage_price_index`.

Overall, the main purpose of this step was to convert the original Excel workbooks into consistent long-format datasets that could be compared more easily across states and time. This also made it easier to carry out the next stage of the project, which involved checking the structure, identifying missing values, reviewing inconsistencies, and cleaning the data where necessary. The tidy datasets and all data processing code used in this project are available in my [GitHub repository](https://github.com/ruby910113/FIT5147-DEP-report).

```{r}
# File path for the overseas migration workbook
om_file_path <- "data/34070DO001_202425.xlsx"

# Sheet names, state names, and row ranges
om_sheet_info <- tibble(
  sheet = paste0("Table 1.", 1:9),
  state = c("Australia", "New South Wales", "Victoria", "Queensland", "South Australia", "Western Australia", "Tasmania", "Northern Territory", "Australian Capital Territory"),
  range = c("A14:W262", rep("A13:W261", 8)))

# Read one overseas migration sheet and tidy it
read_om_sheet <- function(sheet, state, range, file_path) {
  
  om_raw <- read_excel(path = file_path, sheet = sheet, range = range, col_names = FALSE)
  
  # Use the first row as column names
  om_col_names <- om_raw[1, ] %>%
    unlist() %>%
    as.character()
  
  om_data <- om_raw[-1, ]
  names(om_data) <- om_col_names
  
  # Rename the first two columns
  names(om_data)[1:2] <- c("sacc_code", "country_of_birth")
  
  # Remove notes (such as (d)) from the year columns
  om_year_cols <- names(om_data)[-(1:2)]
  om_clean_years <- str_remove(om_year_cols, "\\([A-Za-z]+\\)$")
  names(om_data)[-(1:2)] <- om_clean_years
  
  om_data %>%
    mutate(state = state,
           sacc_code = str_pad(as.character(sacc_code), width = 4, pad = "0")) %>%
    pivot_longer(cols = matches("^\\d{4}-\\d{2}$"),
                 names_to = "financial_year",
                 values_to = "net_overseas_migration") %>%
    mutate(net_overseas_migration = as.numeric(net_overseas_migration)) %>%
    select(state, sacc_code, country_of_birth, financial_year, net_overseas_migration)
}

# Combine all sheets into one dataset
overseas_migration <- pmap_dfr(om_sheet_info, \(sheet, state, range) read_om_sheet(sheet, state, range, om_file_path)) %>%
  filter(!is.na(country_of_birth))

rm(om_sheet_info)
```

```{r}
# Read one housing finance workbook from Data1
read_housing_finance_data1 <- function(file_path, loan_type) {
  
  # Read the metadata rows
  hf_meta <- read_excel(path = file_path, sheet = "Data1", n_max = 10, col_names = FALSE)
  
  # Read the time series data
  hf_data <- read_excel(path = file_path, sheet = "Data1", skip = 10, col_names = FALSE, col_types = c("date", rep("numeric", ncol(hf_meta) - 1)))
  
  # Create temporary column names
  series_cols <- paste0("series_", seq_len(ncol(hf_data) - 1))
  
  names(hf_meta) <- c("label", series_cols)
  names(hf_data) <- c("date", series_cols)
  
  # Make a table for the metadata
  hf_series_info <- tibble(
    series_col = series_cols,
    full_name = as.character(unlist(hf_meta[1, -1])),
    unit = as.character(unlist(hf_meta[2, -1])),
    series_type = as.character(unlist(hf_meta[3, -1])),
    data_type = as.character(unlist(hf_meta[4, -1])),
    frequency = as.character(unlist(hf_meta[5, -1]))) %>%
    separate(full_name,
             into = c("sector", "collection", "borrower_type", "dwelling_type", "state", "measure", "value_type", "blank"),
             sep = "\\s*;\\s*",
             extra = "drop",
             fill = "right") %>%
    select(series_col, state, borrower_type, dwelling_type, measure, value_type, unit, series_type)
  
  # Turn the data into long format
  hf_long <- hf_data %>%
    mutate(year = year(date),
           month = month(date),
           financial_year = if_else(
             month(date) >= 7,
             paste0(year(date), "-", str_sub(year(date) + 1, 3, 4)),
             paste0(year(date) - 1, "-", str_sub(year(date), 3, 4)))) %>%
    pivot_longer(cols = starts_with("series_"),
                 names_to = "series_col",
                 values_to = "housing_finance") %>%
    left_join(hf_series_info, by = "series_col") %>%
    mutate(loan_type = loan_type) %>%
    select(year, month, financial_year, state, loan_type, value_type, unit, series_type, housing_finance)
  
  hf_long
}

# Combine owner-occupier and investor data
housing_finance <- bind_rows(
  read_housing_finance_data1(
    file_path = "data/560104.xlsx", loan_type = "owner_occupier"),
  read_housing_finance_data1(
    file_path = "data/560114.xlsx", loan_type = "investor"))
```

```{r}
# File path for the wage price index workbook
wpi_file_path <- "data/634502b.xlsx"

# Read the metadata rows
wpi_meta <- read_excel(path = wpi_file_path, sheet = "Data1", n_max = 10, col_names = FALSE)

# Read the actual data
wpi_data <- read_excel(path = wpi_file_path, sheet = "Data1", skip = 10, col_names = FALSE, col_types = c("date", rep("numeric", ncol(wpi_meta) - 1)))

# Create temporary column names
series_cols <- paste0("series_", seq_len(ncol(wpi_data) - 1))

names(wpi_meta) <- c("label", series_cols)
names(wpi_data) <- c("date", series_cols)

# Make a metadata table
wpi_series_info <- tibble(
  series_col = series_cols,
  full_name = as.character(unlist(wpi_meta[1, -1])),
  unit = as.character(unlist(wpi_meta[2, -1]))) %>%
  separate(full_name, into = c("wpi_measure", "pay_type", "state", "sector", "industry", "blank"), sep = "\\s*;\\s*", extra = "drop", fill = "right") %>%
  mutate(
    across(c(wpi_measure, pay_type, state, sector, industry), str_squish),
    wpi_measure = case_when(
      wpi_measure == "Quarterly Index" ~ "index",
      wpi_measure == "Percentage Change from Previous Quarter" ~ "pct_change_prev_quarter",
      wpi_measure == "Percentage Change from Corresponding Quarter of Previous Year" ~ "pct_change_prev_year",
      TRUE ~ wpi_measure)) %>%
  select(series_col, state, wpi_measure, unit)

# Turn the data into long format
wage_price_index <- wpi_data %>%
  mutate(year = year(date), month = month(date), 
         financial_year = if_else(
           month(date) >= 7,
           paste0(year(date), "-", str_sub(year(date) + 1, 3, 4)),
           paste0(year(date) - 1, "-", str_sub(year(date), 3, 4)))) %>%
  pivot_longer(cols = starts_with("series_"), names_to = "series_col", values_to = "wage_price_index") %>%
  left_join(wpi_series_info, by = "series_col") %>%
  select(year, month, financial_year, state, wpi_measure, unit, wage_price_index)

rm(wpi_data, wpi_meta, wpi_series_info)
```

## Creating a Common Financial Year Structure

As shown in @tbl-time-summary, Housing finance and Wage price index are quarterly datasets recorded using calendar year and month, while Overseas migration is recorded by financial year. To make the three datasets easier to compare, I created a financial year variable for the quarterly datasets based on the quarter-ending months of September, December, March, and June. The datasets also have different time ranges, since Housing finance and Wage price index start earlier than Overseas migration. Therefore, I restricted all three datasets to the shared financial-year period from 2004-05 to 2024-25 before carrying out the analysis.

```{r}
#| label: tbl-time-summary
#| tbl-cap: Summary of time structure across the datasets

data_time_summary <- bind_rows(
  tibble(
    dataset = "Overseas migration",
    start_ym = "Not applicable",
    end_ym = "Not applicable",
    frequency = "Annual",
    time_period = paste0(min(overseas_migration$financial_year), " to ", max(overseas_migration$financial_year))),
  
  housing_finance %>%
    distinct(year, month, financial_year) %>%
    mutate(date = make_date(year, month, 1)) %>%
    summarise(
      dataset = "Housing finance",
      start_ym = paste0(year(min(date)), "/", month(min(date))),
      end_ym = paste0(year(max(date)), "/", month(max(date))),
      frequency = "Quarterly",
      time_period = paste0(min(financial_year), " to ", max(financial_year))),
  
  wage_price_index %>%
    distinct(year, month, financial_year) %>%
    mutate(date = make_date(year, month, 1)) %>%
    summarise(
      dataset = "Wage price index",
      start_ym = paste0(year(min(date)), "/", month(min(date))),
      end_ym = paste0(year(max(date)), "/", month(max(date))),
      frequency = "Quarterly",
      time_period = paste0(min(financial_year), " to ", max(financial_year))))

data_time_summary %>%
  kable(caption = "Summary of time structure across the datasets",
        col.names = c("Dataset", "Start (Year/Month)", "End (Year/Month)", "Frequency", "Time period")) %>%
  kable_styling(full_width = FALSE)
```

```{r}
overseas_migration <- overseas_migration %>%
  filter(financial_year >= "2004-05",
         financial_year <= "2024-25")

housing_finance <- housing_finance %>%
  filter(financial_year >= "2004-05",
         financial_year <= "2024-25")

wage_price_index <- wage_price_index %>%
  filter(financial_year >= "2004-05",
         financial_year <= "2024-25")
```

## Understanding Missing Values in the Housing Finance Data

@tbl-hf-missing-summary shows that the housing finance dataset is largely complete, except for the Investor Number series, which is only available from 2019-20 onward. This suggests that the missing values are structural rather than accidental, since the series begins later in the original source data. Therefore, these values were not imputed, and any analysis using Investor Number should be limited to 2019-20 onward or interpreted cautiously when compared with the longer series.

```{r}
#| label: tbl-hf-missing-summary

hf_missing_summary <- housing_finance %>%
  group_by(loan_type, value_type) %>%
  summarise(
    first_available = min(financial_year[!is.na(housing_finance)]),
    last_available = max(financial_year[!is.na(housing_finance)]),
    n_missing = sum(is.na(housing_finance)),
    n_available = sum(!is.na(housing_finance)),
    .groups = "drop") %>%
  mutate(
    loan_type = recode(
      loan_type,
      owner_occupier = "Owner-occupier",
      investor = "Investor"))

hf_missing_summary %>%
  kable(caption = "Series start and end dates in the original housing finance workbooks", col.names = c("Loan type", "Value type", "First FY", "Last FY", "Missing values", "Available values")) %>%
  kable_styling(full_width = FALSE)
```

## Checking Distribution Shapes Across the Datasets

@fig-distribution shows no clear sign of major data errors after the asinh transformation. The housing finance data are right-skewed, but the distribution is still fairly smooth, so this looks more like normal variation than unusual mistakes. The overseas migration data are strongly concentrated around zero, with longer tails on both sides, which suggests that many observations are small or zero while some are much larger in either the positive or negative direction. The wage price index data show a clear bimodal shape, but this is likely due to combining index values and percentage change measures in the same dataset, rather than indicating abnormal observations. Overall, this check suggests that the main issue is differences in data structure across the datasets, not obvious errors in the values.

```{r}
#| label: fig-distribution
#| fig-height: 6
#| fig-width: 8

distribution_data <- bind_rows(
  overseas_migration %>%
    transmute(
      dataset = "Overseas migration",
      value = net_overseas_migration),
  housing_finance %>%
    transmute(
      dataset = "Housing finance",
      value = housing_finance),
  wage_price_index %>%
    transmute(
      dataset = "Wage price index",
      value = wage_price_index)) %>%
  filter(!is.na(value)) %>%
  mutate(
    value_asinh = asinh(value),
    dataset = factor(
      dataset,
      levels = c("Housing finance", "Overseas migration", "Wage price index")))

ggplot(distribution_data, aes(x = value_asinh, fill = dataset)) +
  geom_density(alpha = 0.6, colour = "black", linewidth = 0.3) +
  facet_wrap(~ dataset, ncol = 1, scales = "free") +
  labs(
    title = "Distributions of asinh-transformed values across the three datasets",
    x = "asinh(value)",
    y = "Density") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```

## Summary of Data Checking

Overall, the data wrangling and checking steps suggest that the three datasets were successfully converted into tidy formats and aligned to a common financial-year structure for comparison. No major data quality problems were identified, aside from the structurally missing Investor Number series and the expected differences in measurement scales across the datasets. These issues were taken into account in later analysis, so the cleaned data provide a suitable basis for exploring relationships between migration, housing lending, and wage growth across Australian states.

# Data Exploration

This section explores the **relationships between housing lending, overseas migration, and wage growth across Australian states** using financial-year summaries and composition measures from **2004-05 to 2024-25**.

## Does the growth of housing lending align with wage growth across Australian states, and how has the lending structure changed over time?

To address the first research question, I compared annual housing lending growth with annual wage growth across Australian states from 2005-06 to 2024-25. Housing finance was aggregated to annual state-level totals using the Original Value series, while wage growth was calculated from the state-level Wage Price Index. I then examined whether housing lending and wage growth moved together over time, and how the investor share of total lending value changed across states.

```{r}
#| label: q1-prepare-data
#| include: false

state_levels <- c(
  "New South Wales",
  "Victoria",
  "Queensland",
  "South Australia",
  "Western Australia",
  "Tasmania",
  "Northern Territory",
  "Australian Capital Territory")

state_colors <- c(
  "New South Wales" = "#1b9e77",
  "Victoria" = "#d95f02",
  "Queensland" = "#7570b3",
  "South Australia" = "#e7298a",
  "Western Australia" = "#66a61e",
  "Tasmania" = "#e6ab02",
  "Northern Territory" = "#a6761d",
  "Australian Capital Territory" = "#666666")

# Wage growth: state-level and national-level
wpi_state_fy <- wage_price_index %>%
  filter(state %in% state_levels, wpi_measure == "index") %>%
  group_by(financial_year, state) %>%
  summarise(annual_wpi = mean(wage_price_index, na.rm = TRUE), .groups = "drop") %>%
  arrange(state, financial_year) %>%
  group_by(state) %>%
  mutate(wage_growth_pct = (annual_wpi / lag(annual_wpi) - 1) * 100, start_year = as.integer(str_sub(financial_year, 1, 4))) %>%
  ungroup()

wpi_nat_fy <- wage_price_index %>%
  filter(state == "Australia", wpi_measure == "index") %>%
  group_by(financial_year) %>%
  summarise(annual_wpi = mean(wage_price_index, na.rm = TRUE), .groups = "drop") %>%
  arrange(financial_year) %>%
  mutate(wage_growth_pct = (annual_wpi / lag(annual_wpi) - 1) * 100, start_year = as.integer(str_sub(financial_year, 1, 4)))

# Housing lending growth: state-level and national-level
hf_state_fy <- housing_finance %>%
  filter(state %in% state_levels, value_type == "Value", series_type == "Original") %>%
  group_by(financial_year, state) %>%
  summarise(total_lending = sum(housing_finance, na.rm = TRUE), .groups = "drop") %>%
  arrange(state, financial_year) %>%
  group_by(state) %>%
  mutate(lending_growth_pct = (total_lending / lag(total_lending) - 1) * 100, start_year = as.integer(str_sub(financial_year, 1, 4))) %>%
  ungroup()

hf_nat_fy <- housing_finance %>%
  filter(state == "Total Australia", value_type == "Value", series_type == "Original") %>%
  group_by(financial_year) %>%
  summarise(total_lending = sum(housing_finance, na.rm = TRUE), .groups = "drop") %>%
  arrange(financial_year) %>%
  mutate(lending_growth_pct = (total_lending / lag(total_lending) - 1) * 100, start_year = as.integer(str_sub(financial_year, 1, 4)))

# Combined data for Figure 1
overall_growth_states <- bind_rows(
  wpi_state_fy %>%
    transmute(financial_year, start_year, state, series = "Wage growth", growth_pct = wage_growth_pct),
  hf_state_fy %>%
    transmute(financial_year, start_year, state, series = "Housing lending growth", growth_pct = lending_growth_pct)) %>%
  filter(!is.na(growth_pct))

overall_growth_nat <- bind_rows(
  wpi_nat_fy %>%
    transmute(financial_year, start_year, series = "Wage growth", growth_pct = wage_growth_pct),
  hf_nat_fy %>%
    transmute(financial_year, start_year, series = "Housing lending growth", growth_pct = lending_growth_pct)) %>%
  filter(!is.na(growth_pct))

# Investor share of total housing lending value
investor_share_value_state <- housing_finance %>%
  filter(state %in% state_levels, series_type == "Original", value_type == "Value") %>%
  group_by(financial_year, state, loan_type) %>%
  summarise(
    annual_total = sum(housing_finance, na.rm = TRUE),
    .groups = "drop") %>%
  group_by(financial_year, state) %>%
  mutate(share = annual_total / sum(annual_total)) %>%
  ungroup() %>%
  filter(loan_type == "investor") %>%
  mutate(start_year = as.integer(str_sub(financial_year, 1, 4)))

investor_share_value_nat <- housing_finance %>%
  filter(state == "Total Australia", series_type == "Original", value_type == "Value") %>%
  group_by(financial_year, loan_type) %>%
  summarise(
    annual_total = sum(housing_finance, na.rm = TRUE),
    .groups = "drop") %>%
  group_by(financial_year) %>%
  mutate(share = annual_total / sum(annual_total)) %>%
  ungroup() %>%
  filter(loan_type == "investor") %>%
  mutate(start_year = as.integer(str_sub(financial_year, 1, 4)))
```

```{r}
#| label: q1-inline-stats
#| include: false

nat_wage <- overall_growth_nat %>%
  filter(series == "Wage growth")

nat_lending <- overall_growth_nat %>%
  filter(series == "Housing lending growth")

value_state_summary <- investor_share_value_state %>%
  group_by(state) %>%
  summarise(
    mean_share = mean(share, na.rm = TRUE),
    peak_share = max(share, na.rm = TRUE),
    latest_share = share[financial_year == "2024-25"][1],
    .groups = "drop")

value_vs_nat <- investor_share_value_state %>%
  select(financial_year, state, share) %>%
  left_join(
    investor_share_value_nat %>%
      select(financial_year, nat_share = share),
    by = "financial_year") %>%
  group_by(state) %>%
  summarise(
    prop_above_nat = mean(share > nat_share, na.rm = TRUE),
    .groups = "drop")
```


### Divergence between housing lending growth and wage growth

@fig-q1-overall-growth separates housing lending growth and wage growth into two panels, making their state-level patterns easier to compare. Wage growth stays within a relatively narrow range of about `r round(min(nat_wage$growth_pct, na.rm = TRUE), 1)`% to `r round(max(nat_wage$growth_pct, na.rm = TRUE), 1)`%, but still shows a longer-run decline from 2005-06 to 2020-21 before rising again towards its period average of `r round(mean(nat_wage$growth_pct, na.rm = TRUE), 1)`%. Western Australia shows the largest swings in wage growth, while Tasmania appears least affected during the COVID-19 period and rebounds strongly afterwards.

By contrast, housing lending growth is much more volatile and differs more across states. The sharp divergence in 2020-21 likely reflects the COVID-19 period, while the large fall in 2022-23 is consistent with a post-pandemic adjustment. Lending fluctuations are especially large in Northern Territory, Western Australia, and New South Wales, whereas South Australia and the Australian Capital Territory are relatively more stable. Overall, the figure suggests that short-run changes in housing finance are driven much more by lending market conditions than by wage growth.

```{r}
#| label: fig-q1-overall-growth
#| fig-cap: Annual housing lending growth and wage growth across Australian states, 2005-06 to 2024-25. Coloured lines show individual states and the dashed black line shows the Total Australia series.
#| fig-width: 10
#| fig-height: 7

p_lending <- ggplot() +
  geom_line(
    data = overall_growth_states %>%
      filter(series == "Housing lending growth"), aes(x = start_year, y = growth_pct, colour = state), linewidth = 0.9, alpha = 0.9) +
  geom_line(
    data = overall_growth_nat %>%
      filter(series == "Housing lending growth"), aes(x = start_year, y = growth_pct), colour = "black", linetype = "dashed", linewidth = 1.1) +
  scale_colour_manual(values = state_colors, name = "State") +
  scale_x_continuous(breaks = seq(2005, 2024, by = 3), labels = function(x) paste0(x, "-", str_sub(x + 1, 3, 4))) +
  labs(title = "Housing lending growth", x = NULL, y = "Annual growth rate (%)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), plot.title = element_text(face = "bold", hjust = 0.5))

p_wage <- ggplot() +
  geom_line(data = overall_growth_states %>%
              filter(series == "Wage growth"),
            aes(x = start_year, y = growth_pct, colour = state), linewidth = 0.9, alpha = 0.9) +
  geom_line(
    data = overall_growth_nat %>%
      filter(series == "Wage growth"), aes(x = start_year, y = growth_pct), colour = "black", linetype = "dashed", linewidth = 1.1) +
  scale_colour_manual(values = state_colors, name = "State") +
  scale_x_continuous(breaks = seq(2005, 2024, by = 3), labels = function(x) paste0(x, "-", str_sub(x + 1, 3, 4))) +
  labs(title = "Wage growth", x = "Financial year", y = "Annual growth rate (%)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), plot.title = element_text(face = "bold", hjust = 0.5))

(p_lending / p_wage) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Annual Housing Lending Growth and Wage Growth Across Australian States", theme = theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 18))) &
  theme(legend.position = "bottom", legend.title = element_text(size = 11), legend.text = element_text(size = 10))
```

### Long-run changes in the investor share of lending value

@fig-q1-structure-value examines how the investor share of total housing lending value changed across states from 2004-05 to 2024-25. The national pattern is clearly cyclical. The investor share starts at about `r scales::percent(investor_share_value_nat$share[investor_share_value_nat$financial_year == "2004-05"], accuracy = 0.1)`, rises to a peak of `r scales::percent(max(investor_share_value_nat$share, na.rm = TRUE), accuracy = 0.1)` in 2014-15, then falls sharply to `r scales::percent(min(investor_share_value_nat$share, na.rm = TRUE), accuracy = 0.1)` in 2020-21 before recovering to `r scales::percent(investor_share_value_nat$share[investor_share_value_nat$financial_year == "2024-25"], accuracy = 0.1)` by 2024-25.

New South Wales stands out as the most investor-oriented market over the long run. Its average investor share is about `r scales::percent(value_state_summary$mean_share[value_state_summary$state == "New South Wales"], accuracy = 0.1)`, and it remains above the national share in `r scales::percent(value_vs_nat$prop_above_nat[value_vs_nat$state == "New South Wales"], accuracy = 1)` of years, peaking at `r scales::percent(value_state_summary$peak_share[value_state_summary$state == "New South Wales"], accuracy = 0.1)` in 2014-15. By contrast, Tasmania has one of the lowest average investor shares, at about `r scales::percent(value_state_summary$mean_share[value_state_summary$state == "Tasmania"], accuracy = 0.1)`. Northern Territory and Tasmania show a somewhat similar long-run pattern, with investor shares generally remaining below the national benchmark for much of the period. However, both rise sharply after the COVID-19 period. By 2024-25, Northern Territory reaches `r scales::percent(investor_share_value_state$share[investor_share_value_state$state == "Northern Territory" & investor_share_value_state$financial_year == "2024-25"], accuracy = 0.1)` and Tasmania reaches `r scales::percent(investor_share_value_state$share[investor_share_value_state$state == "Tasmania" & investor_share_value_state$financial_year == "2024-25"], accuracy = 0.1)`, compared with the national share of `r scales::percent(investor_share_value_nat$share[investor_share_value_nat$financial_year == "2024-25"], accuracy = 0.1)`. Western Australia follows a different pattern: after 2008-09, its investor share begins to move further away from the national benchmark, before recovering again after the 2020-21 low. It rises from `r scales::percent(investor_share_value_state$share[investor_share_value_state$state == "Western Australia" & investor_share_value_state$financial_year == "2020-21"], accuracy = 0.1)` in 2020-21 to `r scales::percent(investor_share_value_state$share[investor_share_value_state$state == "Western Australia" & investor_share_value_state$financial_year == "2024-25"], accuracy = 0.1)` in 2024-25. The other states generally follow the national cycle more closely and do not show such distinctive departures. Overall, the figure suggests that the structure of housing lending has changed substantially over time and that investor activity has been much more important in some state housing markets than in others.

```{r}
#| label: fig-q1-structure-value
#| fig-cap: Investor share of total housing lending value across Australian states, 2004-05 to 2024-25. Coloured lines show individual states and the dashed black line shows the Total Australia share.
#| fig-height: 5
#| fig-width: 10

ggplot(investor_share_value_state, aes(x = start_year, y = share, colour = state)) +
  geom_line(linewidth = 0.9) +
  geom_line(data = investor_share_value_nat, aes(x = start_year, y = share), inherit.aes = FALSE, colour = "black", linetype = "dashed", linewidth = 1) +
  scale_colour_manual(values = state_colors, name = "State") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_continuous(breaks = seq(2004, 2024, by = 4), labels = function(x) paste0(x, "-", str_sub(x + 1, 3, 4))) +
  labs(title = "Investor Share of Total Housing Lending Value by State", x = "Financial year", y = "Investor share of total lending value") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold", hjust = 0.5, size = 18), legend.position = "bottom", legend.title = element_text(size = 11), legend.text = element_text(size = 10))
```

### Summary of findings

Overall, @fig-q1-overall-growth and @fig-q1-structure-value suggest that housing lending growth does not closely track wage growth across Australian states, and that the structure of lending has changed markedly over time. Investor activity appears to be more strongly related to the composition of lending than to short-run movements in total lending growth, with some states remaining much more investor-oriented than others.

# Conclusion

# Reflection

# References
